---
title: "Analysis on Storms and other severe weather events in public health and economic problems for communities and municipalities in US Storms."
author: "Pradeep Nandi"
date: "8/30/2019"
output: html_document
---

Github repository for the Course: [Reproducible Research](https://github.com/pradeepnandi25/RepData_PeerAssessment2)
</br>
Github repository for Rest of Specialization: [Data Science Specialization](https://github.com/pradeepnandi25/)

## 1: Synopsis
The goal of the assignment is to explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) Storm Database and explore the effects of severe weather events on both population and economy. The events in the database start in the year 1950 and end in November 2011.

The following analysis investigates which types of severe weather events are most harmful on:

1. Health (Injuries and Fatalities) 
2. Property damage and crops (Economic Consequences)

Source of information on the data: [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

## 2: Data Processing.

### 2.1: Data Loading into R Memory.

Download the raw data file and extract the data into a data frame. Then convert raw data into 'data.table'.
```{r DataLoading}
# Load the library's into R Studio memory.
library("data.table")
library("ggplot2")

# Download and read the CSV files into stromDateFrame.
sourceFileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(sourceFileUrl, destfile = paste0(getwd(), './repdata_data_StormData.csv.bz2'))
stromDateFrame <- read.csv("./repdata_data_StormData.csv.bz2")

# Converting data.frame to data.table
stromDateTable <- as.data.table(stromDateFrame)
```

### 2.2: Examining/verify Column Names.

```{r ColumnNames}
# Display list of column's for the data.table (stromDateTable).
colnames(stromDateTable)
```

### 2.3: Data Subsetting.

Subset the dataset on the parameters of interest. Basically, we remove the columns we don't need for clarity. 
```{r DataSubsetting, results = "hide"}
# Finding columns to remove
cols_to_Remove <- colnames(stromDateTable[, !c("EVTYPE"
                                               , "FATALITIES"
                                               , "INJURIES"
                                               , "PROPDMG"
                                               , "PROPDMGEXP"
                                               , "CROPDMG"
                                               , "CROPDMGEXP")])

# Removing columns
stromDateTable[, c(cols_to_Remove) := NULL]

# Only use data where fatalities or injuries occurred.  
stromDateTable <- stromDateTable[(EVTYPE != "?" & 
                                          (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c("EVTYPE"
                                                                                                          , "FATALITIES"
                                                                                                          , "INJURIES"
                                                                                                          , "PROPDMG"
                                                                                                          , "PROPDMGEXP"
                                                                                                          , "CROPDMG"
                                                                                                          , "CROPDMGEXP") ]
```

### 2.4: Converting Exponent Columns into Actual Exponents instead of (-,+, H, K, etc).

Making the PROPDMGEXP and CROPDMGEXP columns cleaner so they can be used to calculate property and crop cost.
```{r CorrectingExponents, results = "hide"}

# Change all damage exponents to uppercase.
cols <- c("PROPDMGEXP", "CROPDMGEXP")
stromDateTable[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]

# Map property damage alphanumeric exponents to numeric values.
propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)

# Map crop damage alphanumeric exponents to numeric values
cropDmgKey <-  c("\"\"" = 10^0,
                 "?" = 10^0, 
                 "0" = 10^0,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)

stromDateTable[, PROPDMGEXP := propDmgKey[as.character(stromDateTable[, PROPDMGEXP])]]
stromDateTable[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]

stromDateTable[, CROPDMGEXP := cropDmgKey[as.character(stromDateTable[, CROPDMGEXP])] ]
stromDateTable[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]
```

### 2.5: Making Economic Cost Columns.

```{r EconomicCostColumns}
stromDateTable <- stromDateTable[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propCost = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]
```

### 2.6: Calcuating Total Property and Crop Cost.

```{r TotalPropertyCropCost}
totalCostDataTable <- stromDateTable[, .(propCost = sum(propCost), cropCost = sum(cropCost), Total_Cost = sum(propCost) + sum(cropCost)), by = .(EVTYPE)]

totalCostDataTable <- totalCostDataTable[order(-Total_Cost), ]

totalCostDataTable <- totalCostDataTable[1:10, ]

head(totalCostDataTable, 5)
```

### 2.7: Calcuating Total Fatalities and Injuries.

```{r TotalFatalitiesInjuriesCalc}
totalInjuriesDataTable <- stromDateTable[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]

totalInjuriesDataTable <- totalInjuriesDataTable[order(-FATALITIES), ]

totalInjuriesDataTable <- totalInjuriesDataTable[1:10, ]

head(totalInjuriesDataTable, 5)
```

## 3: Results.

### 3.1: Events that are Most Harmful to Population Health

Melting data.table so that it is easier to put in bar graph format 
```{r HealthResults}
bad_stuff <- melt(totalInjuriesDataTable, id.vars="EVTYPE", variable.name = "bad_thing")
head(bad_stuff, 5)
```

```{r healthChart}
# Create chart
healthChart <- ggplot(bad_stuff, aes(x = reorder(EVTYPE, -value), y = value))

# Plot data as bar chart
healthChart = healthChart + geom_bar(stat = "identity", aes(fill = bad_thing), position = "dodge")

# Format y-axis scale and set y-axis label
healthChart = healthChart + ylab("Frequency Count") 

# Set x-axis label
healthChart = healthChart + xlab("Event Type") 

# Rotate x-axis tick labels 
healthChart = healthChart + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Set chart title and center it
healthChart = healthChart + ggtitle("Top 10 US Killers") + theme(plot.title = element_text(hjust = 0.5))

healthChart
```

### 3.2: Events that have the Greatest Economic Consequences

Melting data.table so that it is easier to put in bar graph format 
```{r EconConsequences}
econ_consequences <- melt(totalCostDataTable, id.vars="EVTYPE", variable.name = "Damage_Type")
head(econ_consequences, 5)
```

```{r econChart}
# Create chart
econChart <- ggplot(econ_consequences, aes(x = reorder(EVTYPE, -value), y = value))

# Plot data as bar chart
econChart = econChart + geom_bar(stat = "identity", aes(fill = Damage_Type), position = "dodge")

# Format y-axis scale and set y-axis label
econChart = econChart + ylab("Cost (dollars)") 

# Set x-axis label
econChart = econChart + xlab("Event Type") 

# Rotate x-axis tick labels 
econChart = econChart + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Set chart title and center it
econChart = econChart + ggtitle("Top 10 US Storm Events causing Economic Consequences") + theme(plot.title = element_text(hjust = 0.5))

econChart
```